---
title: "STAT 630 - Final Project"
author: "Shrusthi, Pranjal, Jeevan"
format: pdf
editor: visual
---

# 1. Setup & Loading the datset

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)


library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
student <- read_csv("/Users/pranjalts/Desktop/CSUEB/sem_1/SM/DAP/student-performance-analysis/student_maths.csv")
```

# 2. Data Import & Cleaning

let's convert the ‘paid’ column into a factor so R treats it as a categorical grouping variable for the two-means hypothesis test.

```{r}
student <- student |>
select(G3, paid, studytime, failures, absences, Medu, Fedu) |>
mutate(
paid = as.factor(paid)
)
```

```{r}
table(student$paid)
```

```{r}
colSums(is.na(student))
```

There are no missing values in the dataset.

```{r}
summary(student)
```

Final grades (G3) range from 0–20. Studytime values (1–4) represent weekly study hours from less than 2 hours to more than 10 hours. The ‘failures’ variable ranges from 0–3, where it represents past failures. Absences range from 0 to 75, indicating the total number of school days a student missed. Parental education levels (Medu and Fedu) range from 0–4, representing increasing levels of education. The paid tutoring groups include 214 of ‘no’ and 181 of ‘yes’ students.

# 3. Exploratory Data Analysis

```{r}
student |>
  group_by(paid) |>
  summarise(
    mean = mean(G3),
    sd = sd(G3),
    median = median(G3),
    n = n()
  )
```

```{r}
student |> 
  ggplot(aes(G3, fill = paid)) +
  geom_histogram(binwidth = 2,
                 col = "white",
                 show.legend = FALSE) +
  facet_wrap(~ paid) +
  labs(title = "Distribution of Final Math Grades by Paid Tutoring",
       x = "Final Math Grade (G3)",
       y = "Count")

```

```{r}
student |>
  ggplot(aes(x = absences, y = G3)) +
  geom_point(alpha = 0.6) +
  labs(title = "Absences vs Final Grade (G3)",
       x = "Number of Absences",
       y = "Final Grade")
```

```{r}

student |>
  ggplot(aes(x = as.factor(studytime), y = G3, fill = as.factor(studytime))) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Final Grade (G3) by Weekly Study Time",
       x = "Weekly Study Time (1 to 4)",
       y = "Final Grade (G3)")

```

```{r}
student |>
  ggplot(aes(x = as.factor(failures), fill = paid)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Past Failures by Paid Tutoring Group",
       x = "Number of Past Failures",
       y = "Count",
       fill = "Paid Tutoring")

```

```{r}
student <- student |>
  mutate(parent_edu_group = as.factor(round((Medu + Fedu) / 2)))

student |>
  ggplot(aes(x = parent_edu_group, y = G3, fill = parent_edu_group)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Final Grades by Average Parental Education Level",
       x = "Average Parental Education (0 to 4)",
       y = "Final Grade (G3)")

```

```{r}
student <- student |>
  mutate(parent_edu_group = as.factor(round((Medu + Fedu) / 2)))

student |>
  ggplot(aes(x = parent_edu_group, fill = paid)) +
  geom_bar(position = "dodge") +
  labs(title = "Parental Education Level vs Paid Tutoring",
       x = "Average Parental Education (0 to 4)",
       y = "Count of Students",
       fill = "Paid Tutoring")

```

# 3. Assumption Checks:

-   Hypotheses:

-   **Null hypothesis (H0):** The population mean final math grades are equal for students who receive paid tutoring and those who do not.

    $H_0: \mu_1 = \mu_2$

-   **Alternative hypothesis (H A):** The mean final math grade for students who receive paid tutoring is greater than that of students who do not.

    $H_A: \mu_1 > \mu_2$

Where,

$\mu_1$ = the population mean final math grade for students who receive paid tutoring.

$\mu_2$ = the population mean final math grade for students who do not receive paid tutoring.

-   Check conditions:

*Independence* - Each observation represents a different student, so the observations within each group are independent. The two groups paid tutoring and no paid tutoring do not overlap, so the groups are independent of each other.

*Normality*

```{r}
paid_yes <- student |>
  filter(paid == "yes") |>
  pull(G3)

paid_no <- student |>
  filter(paid == "no") |>
  pull(G3)

n1 <- length(paid_yes)
n2 <- length(paid_no)

n1; n2
```

Since both sample sizes exceed 30, the Central Limit Theorem supports approximate normality of the sampling distribution for the mean in each group.

```{r}
student |>
  ggplot(aes(sample = G3)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ paid) +
  labs(title = "Q–Q Plots of Final Grades by Paid Tutoring",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")
```

# 4.Statistical Analysis:

-   test statistic

```{r}
xbar1 <- mean(paid_yes)
xbar2 <- mean(paid_no)

s1 <- sd(paid_yes)
s2 <- sd(paid_no)

stat <- xbar1 - xbar2
null_value <- 0
se <- sqrt(s1^2/n1 + s2^2/n2)

df <- min(c(n1, n2)) - 1

t_stat <- (stat - null_value) / se
t_stat
```

-   p-value

```{r}
p_val <- pt(t_stat, df = df, lower.tail = FALSE)
p_val
```

-   decision and conclusion:

Decision: reject $H_0$

We have enough evidence that the true difference of the mean final math grade (G3) for students who receive paid tutoring compared to students who do not receive paid tutoring is **more than 0**.

-   95% Confidence Interval

```{r}
t.test(paid_yes, paid_no,
       alternative = "greater",
       conf.level = 0.95)
```
